Wait-free Queues with Polylogarithmic Step Complexity

### 2022-10-05 thesis/thesis.tex
add figures, fix bugs, minor modifications

\If{\tt{dir {\keywordfont is} right}} \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-right}}  \label{considerPreviousRight}
\State \tt{i+= blocks[sb].num\sub{deq-left}}  \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDequeue}{sb, i}
\EndIf
\EndFunction{IndexDequeue}

@@

$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads value $b$ for $n.$\nf{head}. Since $R_{p2}$'s \nf{CAS} into $n$\nf{parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} after $R_{p1}$'s \ref{cas} and before $R_{p2}$'s \ref{readHead}. We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$. Finally, since \ref{readChildHead} is after \ref{readHead} and $R_{p2}$'s \ref{readHead} is after $R_p^\prime$'s \ref{incrementHead}, $R_{p2}$ reads a value $\geq b$ for $n.\nf{head}$.

$$_{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_p1}<_{\ref{incrementHead}}^{R_p^\prime}t<_{\ref{cas}}^{R_p2}t<_{\ref{readChildHead}}^{R_p2}$$
TODO: draw figure

@@

Furthermore, if $B$\nf{.super} is not set earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.

@@

\begin{lemma}[\nf{IndexDequeue} correctness]
 If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
\end{lemma}
\begin{proof}
We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $sb$, index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}. After that the position of $D_i(n,b)$ in $D(n.\nf{parent}, sb)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}. By Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right. If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft}. Else if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}). Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis. 
\end{proof}
TODO: add figure

>>>>

\If{\tt{dir {\keywordfont is} right}} \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-right}}  \label{considerPreviousRight}
\State \tt{i+= parent.blocks[sb].num\sub{deq-left}}  \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDequeue}{sb, i}
\EndIf
\EndFunction{IndexDequeue}

@@

$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads value $b$ for $n.$\nf{head}. Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} (Line \ref{incrementHead}) after $R_{p1}$'s \ref{cas} and before $R_{p2}$'s \ref{readHead}. We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}. Finally, since \ref{readChildHead} is after \ref{readHead} and $R_{p2}$'s \ref{readHead} is after $R_p^\prime$'s \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}. So $R_{p2}$ reads a value $\geq b$ for $n.\nf{head}$ by Lemma \ref{nonDecreasingHead}.
\begin{align*}
\left.
\begin{array}{r}
 _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
_{\ref{cas}}^{R_{p1}}<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
\end{array}
\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
\end{align*}

@@

Furthermore, if $B$\nf{.super} was not set earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.

@@

\begin{lemma}[\nf{IndexDequeue} correctness]
 If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
\end{lemma}
\begin{proof}
We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $sb$, index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}. After that the position of $D_i(n,b)$ in $D(n.\nf{parent}, sb)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}. By Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right. If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}). Else if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}). Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis. 
\end{proof}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}

  \caption{The number of \nf{dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a left child.}\label{fig::blockSumDeqLeft}
\end{figure}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}

  \caption{The number of \nf{dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a right child.}\label{fig::blockSumDeqRight}
\end{figure}
